Extreme heterogeneity of influenza virus infection in single cells

### 2018-01-14 paper/ResponseToReviews.tex

The main shortcoming of the paper is that it provides very little in the way of new information. As the authors clearly and honestly point out, most of the findings in this paper are simply confirming observations made in older papers. Additionally, the findings are purely descriptive, and provide little insight into the mechanisms that may give rise to the observations.

We believe that addressing the points below would help ameliorate some of the shortcomings of the paper.

{\color{black}
This is a fair summary of our work.
Thank you for recognizing the power of the approach and our efforts to fully describe pitfalls and effectively present the complex datasets.

As you note, our approach does not uncover any truly new biological phenomenon, and (as you also note) we honestly describe how our results end up mostly confirming older observations made using more indirect methods.
However, there is substantial value to explicitly observing these processes in single cells, even if the findings are mostly in line with existing circumstantial evidence from bulk studies or less comprehensive single-cell approaches (e.g., flow cytometry).
Also, the extremeness of the heterogeneity across cells is surprising even in light of all that was previously known.

We have revised our manuscript in response to the many helpful suggestions from the reviewers.
As you will see, we have addressed all the major questions and suggestions.
}

@@

\section*{MAJOR COMMENTS}

1. There are some obvious sources of variability that haven't been suitably discussed.

- The first is variation in the timing of the early stages of infection. No steps appear to have been taken to synchronize infection, or to limit secondary spread of the virus within the cultures. Could heterogeneity in the timing of binding/entry/fusion/trafficking explain a lot of the variation observed?

{\color{black}
This is an important point.
We did not synchronize infection, although for our 8-hour samples, we washed the cells one hour after infection for one sample (8hr) but not the other (8hr-2) in order to evaluate the importance of infection timing.
The heterogeneity was similar between these two samples, as demonstrated by the new quantification method (Gini coefficients) suggested by the reviewers in a later comment.

We have also performed new experiments where we synchronized infection by pre-binding virus on ice, washing it away, and then raising the temperature.
Analysis of this experiment by flow cytometry staining is in the new Figure 4-Figure supplement 4, and shows that these synchronized infections look similar to ones that are neither synchronized nor washed.

Finally, the barcoded viruses (Figure 4B) provide strong evidence against secondary infection, since secondary infection would tend to lead to mixed-barcode cells because there would be many of progeny viruses of both barcodes.

@@

- As the variability of the viral RNA load appears to be the central result of the manuscript, it would be useful to:

    a. Employ a statistic to quantify the variability (e.g. entropy or Gini index)

    b. Use simple models to illustrate how much variability can be accounted for by simple effects, such as expected Poisson co-infection frequency, or the likelihood of attaining full complement of RNP genes.

{\color{black}
The idea of quantifying the variability in viral mRNA across infected cells is a good one.
We have done this using the Gini coefficient.
The results are in Figure 4--Figure supplement 3, and are referred to in the text related to that figure.
As a fun point of comparison, the Gini coefficients for the variability of viral mRNA across cells ($\ge$0.64 for all samples) exceed the Gini coefficient for the unevenness of income distribution in the United States.
}

@@

The reviewers are correct that our statement about the role of co-infection in influencing cell-to-cell variation was too strong.
We have clarified this statement.
We have also added a figure supplement (Figure 4-Figure supplement 5) which uses an independent method (flow cytometry staining of individual viral proteins) to support the fact that even at high infectious dose both low- and high-producing cells remain, although their relative proportions change.
This figure supplement also supports our clarified statement that viral mRNA (or protein) production is not a simple continuous function of infectious dose, but is also affected by other sources of heterogeneity.
The new text reads:
\begin{quote}
\textsl{
Notably, Figure 4E shows that there are co-infected cells with both low and high amounts of viral mRNA, suggesting that the initial infectious dose does not drive a simple continuous increase in viral transcript production.
In support of this view, we used flow cytometry to quantify the levels of individual viral proteins in cells infected at various MOIs or for which we could delineate co-infection status (Figure 4-Figure supplement 5).
This analysis shows that sub-populations of cells that express similarly low and high levels of viral proteins persist across a wide range of infectious doses, although co-infection can influence the relative proportion of infected cells that fall into these sub-populations (Figure 4-Figure supplement 5).
}
\end{quote}
}

3. The analysis of host cell transcripts positively or negatively associated with high viral mRNA expression is pretty minimal. Do the host genes identified here match up with the results of previous studies that screened for host pro- and anti-viral factors (reviewed in Watanabe et al. DOI: 10.1016/j.chom.2010.05.008)? Also, targeted gene knockdown or over-expression experiments could help establish the causality underlying these relationships.

{\color{black}
We have substantially expanded the last section of the Results that focuses on cellular gene expression.
We call out the gene-set enrichments, which show that genes involved in oxidative stress responses are at higher levels in cells with more viral mRNA.
The possible pro-viral role of transient oxidative stress has been reported before, and we have performed new experiments (Figure 9-Figure supplement 2) that directly support the idea that it is pro-viral.
These new experiments therefore suggest causality for one of the host-cell factors that we identify.

@@

We also compare the genes that we find significantly associated with viral mRNA to the 128 genes highlighted by Watanabe et al (2010) as having been found multiple times in genome-wide studies.
There is no overlap.
This fact is discussed in the last paragraph of the Results.
We theorize that this is because most of the cell-to-cell heterogeneity during infection of ``normal'' (e.g., not genetically perturbed) cells is \emph{not} due to changes in expression of the type of large-effect genes that are typically identified in genome-wide screens.
Rather, our results would seem to suggest that a substantial amount of the heterogeneity is due to viral / stochastic factors (e.g., failure to express certain viral genes), or perhaps more subtle changes in cellular state such as transient oxidative stress.

Importantly, we do \emph{not} believe that the main value of our approach is as a method to identify pro- or anti-viral genes.
Our goal is to figure out the actual extent and sources of cell-to-cell heterogeneity, which may not be due to primarily alterations in the expression level of single host genes with large effects.
}

@@

We have clarified that we are \emph{only} using Figure 2C to show that the high DI stocks induce more IFN per TCID50, and that this could be due entirely to more physical virions.
This is an important point for readers less versed in how viral passaging / DI content can influence IFN induction, since many studies simply report on the TCID50s used and provide no information on stock purity.
This fact likely explains the discrepancy in IFN induction in our study and prior bulk studies (e.g., \textit{eLife}, 5:e18311) on A549 cells that used viruses passaged at high MOIs that almost certainly contained many DI particles.

The revised text clearly explains how Figure 2C is comparing stocks with equal TCID50 but not equal physical virions. 
Specifically, we have added the following statement to the legend for Figure 2C, and similar sentences in the manuscript main text:
\begin{quote}
\textsl{
Our viral stocks are less immunostimulatory than virus propagated at high MOI when used at the same number of infectious units as calculated by TCID50.
Note that this fact does not necessarily imply that they are more immunostimulatory per virion, as the high-MOI stocks also have more virions per infectious unit as shown in the first two panels (Figure 2A,B).}
\end{quote}
}

- Fig. 3B: While I appreciate the density of information presented by this figure, the authors' main point concerning heterogeneity in the viral proportion of total mRNA reads might be made more clearly by simply plotting the relative proportion of mRNAs that are viral per cell.

{\color{black}
The reviewer points out that Figure 3B is a complicated way to show the heterogeneity in total mRNA.
In response to this comment, we have added Figure 3 - Figure supplement 1, which shows the distribution of the viral proportion as a cumulative fraction plot.

### 2018-01-14 paper/paper.tex
updated to new figure referencing enabled by new _eLife_ LaTex class file, this cleans up the LaTex source a bit

Infected cells will express viral as well as cellular mRNAs -- however the cell barcodes and UMIs cannot distinguish whether a cell was initially infected by one or multiple viral particles.
We therefore engineered an influenza virus (strain A/WSN/1933) that additionally carried \emph{viral barcodes} consisting of synonymous mutations near the 3' end of each transcript (Figure~\ref{fig:workflow}A).
Critically, these synonymous mutations did not greatly impact viral growth kinetics (Figure~\ref{fig:workflow}B).
We infected A549 human lung carcinoma cells with an equal mix of the wild-type and synonymously barcoded viruses.
Cells infected by a single virion will exclusively express mRNAs from either wild-type or synonymously barcoded virus, whereas cells that are co-infected with multiple virions will often express mRNAs from both the wild-type and synonymously barcoded viruses (Figure~\ref{fig:workflow}C).

We took care to generate stocks of virus that were relatively ``pure'' of defective particles.
Stocks of viruses typically contain an array of biologically active viral particles, some of which are defective for replication owing to mutations or deletions in essential viral genes~\citep{von1954incomplete,huang1970defective,Brooke:2014ch,Fonville:2015cg,Lauring:2010if,Dimmock:2014dk,saira2013sequence}.
These defective particles become prevalent when a virus is grown at high MOI, where complementation permits the growth of otherwise deleterious genotypes.
To minimize the levels of defective particles, we propagated our viral stocks at low MOI for a relatively brief period of time~\citep{Xue:2016bw}.
We validated that our stocks exhibited greater purity of infectious particles than a stock propagated at high MOI by verifying that they had a higher ratio of infectious particles to virion RNA (Figure~\ref{fig:viruspopulations}A) and to particles capable of inducing expression of a single viral protein (Figure~\ref{fig:viruspopulations}B).
In addition, viral stocks with many defective particles are more immunostimulatory per infectious unit (e.g., TCID50) than low-defective stocks~\citep{Tapia:2013kf,Lopez:2014en}, in part simply because there are more physical virions per infectious unit (Figure~\ref{fig:viruspopulations}A,B).
We confirmed that our viral stocks induced less interferon per infectious unit than a stock propagated at higher MOI (Figure~\ref{fig:viruspopulations}C).

\begin{figure}
\centerline{\includegraphics[width=0.7\linewidth]{figures/Validating_barcode_virus/validating_populations_D02.pdf}}
\caption{\label{fig:viruspopulations} The viral stocks in our experiments are relatively pure of defective particles. 
{\bf (A)}
Our viral stocks have a higher ratio of infectious particles to HA virion RNA compared to a high-defective stock propagated at high MOI.
HA viral RNA was quantified by qPCR on virions. 
Error bars $\pm$ S.D., n=6 (qPCR replicates). 
{\bf (B)} 
Our viral stocks have a higher ratio of infectious particles to particles capable of expressing HA protein.
A549 cells were infected at an MOI of 0.1, and the percentage of cells expressing HA protein at 9 hours post-infection was quantified by antibody staining and flow cytometry.
{\bf (C)} 
Our viral stocks are less immunostimulatory than virus propagated at high MOI when used at the same number of infectious units as calculated by TCID50.
Note that this fact does not necessarily imply that they are more immunostimulatory per virion, as the high-MOI stocks also have more virions per infectious unit as shown in the first two panels. 
Measurements of \textit{IFNB1} transcript by qPCR normalized to the housekeeping gene L32 in A549 cells at 10 hours post infection at an MOI of 0.5.
Error bars $\pm$ S.D., n=3.
Note that MOIs were calculated by TCID50 on MDCK-SIAT1 cells, whereas the experiments in this figure involved infection of A549 cells.
}
\figsupp[Full flow cytometry data for panel B . \label{figsupp:HAstain}]{
Full flow cytometry data for  Figure~\ref{fig:viruspopulations}B. 
A549 cells were infected at an MOI of 0.1 as calculated by TCID50 on MDCK-SIAT1 cells. 
{\bf (A)} Uninfected gating control.
{\bf (B)} Cells infected with the wild-type virus stock used in our experiments.
{\bf (C)} Cells infected with synonymously barcoded virus stock used in our experiments.
{\bf (D)} Cells infected with a stock of wild-type virus propagated at a high MOI, and therefore enriched in defective particles.
}{\includegraphics[width=0.7\linewidth]{figures/Validating_barcode_virus/HA_stain_supplement.png}}
\end{figure}

@@

We recovered between 3,000 and 4,000 cells for each sample (Figure~\ref{fig:cells}A). 
As expected for a low-MOI infection, most cells expressed little or no viral mRNA (Figure~\ref{fig:cells}B and Figure~\ref{fig:cells}-Figure~supplement~\ref{figsupp:viral_mRNA_cumulfrac}).
Also as expected, the amount of viral mRNA per cell among infected cells increased over time (Figure~\ref{fig:cells}B, Figure~\ref{fig:cells}-Figure~supplement~\ref{figsupp:viral_mRNA_cumulfrac}).
But what was most notable was how widely the number of viral mRNA molecules varied among infected cells.
While the fraction of mRNA derived from virus was $<$0.1\% for most cells, viral mRNA constituted half the transcriptome in a few cells at 8 and 10 hours (Figure~\ref{fig:cells}B, Figure~\ref{fig:cells}-Figure~supplement~\ref{figsupp:viral_mRNA_cumulfrac}).

\begin{figure}
\centerline{\includegraphics[width=\linewidth]{figures/p_cell_mRNA_summary.pdf}}
\caption{\label{fig:cells}
There is a very wide distribution in the amount of viral mRNA per cell.
{\bf (A)} 
Number of cells sequenced for each sample.
{\bf (B)} 
The number of cellular and viral mRNAs detected for each cell is plotted as a point.
The blue lines show the overall distribution of the number of cellular mRNAs per cell.
The orange rug plot at the left of each panel shows the distribution of the number of viral mRNAs per cell.
Cells outside the dotted green lines were considered outliers with suspiciously low or high amounts of cellular mRNA (possibly derived from two cells per droplet), and were excluded from all subsequent analyses.
Figure~\ref{fig:cells}-Figure~supplement~\ref{figsupp:viral_mRNA_cumulfrac} shows the exact distributions of the fraction of viral mRNA per cell.
}
\figsupp[Cumulative fraction plot of proportion of total mRNA from virus. \label{figsupp:viral_mRNA_cumulfrac}]{
For each sample, this plot shows the fraction of all cells that derive at least the indication fraction of their mRNA from influenza virus.
}{\includegraphics[width=0.5\linewidth]{figures/p_mRNA_per_cell_cumul.pdf}}
\end{figure}

A complicating factor is that uninfected cells could have small amounts of viral mRNA due to leakage of transcripts from lysed cells.
It is therefore important to establish a threshold for identifying truly infected cells.
We can do this by taking advantage of the fact that roughly half the infecting virions bear synonymous barcodes.
Reads derived from lysed cells will be drawn from both wild-type and synonymously barcoded viral transcripts.
However, most cells are infected by at most one virion, and so the reads from truly infected cells will usually derive almost entirely from one of the two viral variants. 
Figure~\ref{fig:viralbarcodes}A shows the fraction of viral reads in individual cells from each viral variant, and Figure~\ref{fig:viralbarcodes}B indicates the fraction of viral reads from the most abundant variant in that cell.
Most cells with large amounts of viral mRNA have viral transcripts exclusively derived from one viral variant -- indicating non-random partitioning as expected from viral infection.
However, cells with a small amount of viral mRNA often have viral transcripts from both variants, as expected from the random partitioning associated with simple mRNA leakage.
Finally, a few cells with large amounts of viral mRNA have viral transcripts from both variants, likely reflecting co-infection.

\begin{figure}[t!]
\centerline{\includegraphics[width=0.85\linewidth]{figures/p_frac_flu_summary.pdf}}
\caption{\label{fig:viralbarcodes}
Synonymous barcodes on the viral mRNAs distinguish true infections from cells that contain viral mRNAs derived from leakage of lysed cells.
{\bf (A)}
Cells with at least two viral mRNAs for which the barcode could be called, arranged in order of increasing influenza transcript counts.
Bar heights denote the number viral mRNAs on a log\textsubscript{10} scale, bar coloring is linearly proportional to the fractions of viral mRNAs derived from wild-type and synonymously barcoded virus.
{\bf (B)}
Same as (A), but each bar is colored according to the relative fraction of the more common (major) and less common (minor) virus variant.
At low levels of viral mRNA there is often a roughly equal mix, suggesting contamination with viral mRNAs leaked from lysed cells.
At higher levels of viral mRNA, cells generally have only one viral variant, suggesting infection initiated by a single virion.
A few cells are also obviously co-infected with both viral variants.
{\bf (C)}
We determined a threshold for calling ``true'' infections by finding the amount of viral mRNA per cell at which the viral barcode purity no longer increases with more viral mRNA.
The purity is the fraction of all viral mRNA in a cell derived from the most abundant viral barcode in that cell.
We fit a curve (orange line) to the mean purity of all cells with more than the indicated amount of viral mRNA, and drew the cutoff (dotted green line) at the point where this curve stopped increasing with the fraction of total mRNA derived from virus.
This plot illustrates the process for the 10-hour sample, see Figure~\ref{fig:viralbarcodes}-Figure~supplement~\ref{figsupp:thresholds} for similar plots for other samples.
See the Methods for details.
{\bf (D)}
The number of cells identified as infected and co-infected for each sample, as well as the number of cells with any viral read.
For all subsequent analyses, we subsampled the number of uninfected cells per sample to the greater of 50 or the number of infected cells.
{\bf (E)} 
Distribution of the fraction of mRNA per cell derived from virus for both infected and co-infected cells.
Figure~\ref{fig:viralbarcodes}-Figure~supplement~\ref{figsupp:cumulfracflu} shows these same data in a cumulative fraction plots and calculates Gini coefficients to quantify the heterogeneity in viral mRNA load.
}

@@

\figsupp[Thresholds for calling infected cells. \label{figsupp:thresholds}]{
Cell lysis can lead cells to the spurious association of small amounts of extraneous mRNA with individual cells.
We wanted to avoid classifying as infected cells that had simply acquired such lysis-derived viral mRNA.
The amount of lysis-derived viral mRNA will vary among samples as a function of both the lysis rate during the cell preparation (which always varies slightly from sample to sample in the 10X procedure) and with the amount of total viral mRNA for that sample (the more viral mRNA, the more there is to be acquired from lysed cells).
As is shown in Figure~\ref{fig:viralbarcodes}B, the 8hr-2 and 10hr sample clearly have an enrichment of mixed barcodes in cells with small numbers of viral mRNA.
For each sample, we calculated the mean purity of all cells with at least the indicated amount of viral mRNA, and determined the threshold amount of viral mRNA where purity no longer increased by finding the first maxima in a loess curve fit (orange line).
We called the threshold at this point of maximum purity (dotted green line).
For the 6hr and 8hr samples there is no indication of contamination from lysis-derived reads, as Figure~\ref{fig:viralbarcodes}B shows no increase in mixed barcodes in low viral mRNA cells.
Therefore, for these samples we simply set a threshold of requiring at least 2$\times$10$^{-4}$ of the total mRNA to come from virus, which corresponds to $\sim$2 viral mRNAs for the typical cell with $10^4$ total reads (Figure~\ref{fig:cells}B).
}{\includegraphics[width=\linewidth]{figures/p_purity_threshold.pdf}}

@@

We determined the threshold amount of viral mRNA per cell for each sample at which the barcode partitioning clearly resulted from infection rather than leakage (Figure~\ref{fig:viralbarcodes}C and Figure~\ref{fig:viralbarcodes}-Figure~supplement~\ref{figsupp:thresholds}), and used these thresholds to annotate cells that we were confident were truly infected.
We also annotated as co-infected cells above this threshold that had mRNA from both viral variants.
Figure~\ref{fig:viralbarcodes}D shows the number of cells annotated as infected and co-infected for each sample -- these cells are just a small fraction of the number of cells with any viral read.
These annotation thresholds are conservative, and may miss some true low-level infections.
However, it is important that the analyses below are restricted to cells that are truly infected with virus, so we accepted the possible loss of some low-level infections in order to avoid false positives.
In addition, the synonymous viral barcodes only identify co-infections by viruses with different barcodes -- since the barcodes are at roughly equal proportion, we expect to miss about half of the co-infections.
Since we annotate about $\sim$10\% of the infected cells as co-infected by viruses with different barcodes (Figure~\ref{fig:viralbarcodes}D), we expect another $\sim$10\% of the infected cells to also be co-infected but not annotated as so by our approach.
Because most cells are not infected, we subsampled the uninfected cells to the numbers shown in Figure~\ref{fig:viralbarcodes}D to balance the proportions of infected and uninfected cells for all subsequent analyses.

Strikingly, the extreme variation in the number of viral transcripts per cell remains even after we apply these rigorous criteria for annotating infected cells (Figure~\ref{fig:viralbarcodes}E). 
The fraction of viral mRNA per infected cell follows a roughly exponential distribution, with many cells having few viral transcripts and a few cells having many.
At 6 and 8 hours $<$10\% of infected cells are responsible for over half the viral transcripts, while at 10 hours $<$15\% of infected cells produce over half the viral transcripts (Figure~\ref{fig:viralbarcodes}-Figure~supplement~\ref{figsupp:cumulfracflu}).
One way to quantify the heterogeneity of a distribution is to calculate the Gini coefficient~\citep{gini1921measurement}, which ranges from 0 for a completely uniform distribution, to 1 for a maximally skewed distribution.
Figure~\ref{fig:viralbarcodes}-Figure~supplement~\ref{figsupp:cumulfracflu} shows the Gini coefficients for the distribution of viral mRNA across infected cells for each sample.
The Gini coefficients are $\ge$0.64 for all samples.
As a fun point of comparison, these Gini coefficients indicate that the distribution of viral mRNA across infected cells is more uneven than the distribution of income in the United States~\citep{alvaredo2011note}.

>>>>

Infected cells will express viral as well as cellular mRNAs -- however the cell barcodes and UMIs cannot distinguish whether a cell was initially infected by one or multiple viral particles.
We therefore engineered an influenza virus (strain A/WSN/1933) that additionally carried \emph{viral barcodes} consisting of synonymous mutations near the 3' end of each transcript (\FIG{workflow}A).
Critically, these synonymous mutations did not greatly impact viral growth kinetics (\FIG{workflow}B).
We infected A549 human lung carcinoma cells with an equal mix of the wild-type and synonymously barcoded viruses.
Cells infected by a single virion will exclusively express mRNAs from either wild-type or synonymously barcoded virus, whereas cells that are co-infected with multiple virions will often express mRNAs from both the wild-type and synonymously barcoded viruses (\FIG{workflow}C).

We took care to generate stocks of virus that were relatively ``pure'' of defective particles.
Stocks of viruses typically contain an array of biologically active viral particles, some of which are defective for replication owing to mutations or deletions in essential viral genes~\citep{von1954incomplete,huang1970defective,Brooke:2014ch,Fonville:2015cg,Lauring:2010if,Dimmock:2014dk,saira2013sequence}.
These defective particles become prevalent when a virus is grown at high MOI, where complementation permits the growth of otherwise deleterious genotypes.
To minimize the levels of defective particles, we propagated our viral stocks at low MOI for a relatively brief period of time~\citep{Xue:2016bw}.
We validated that our stocks exhibited greater purity of infectious particles than a stock propagated at high MOI by verifying that they had a higher ratio of infectious particles to virion RNA (\FIG{viruspopulations}A) and to particles capable of inducing expression of a single viral protein (\FIG{viruspopulations}B).
In addition, viral stocks with many defective particles are more immunostimulatory per infectious unit (e.g., TCID50) than low-defective stocks~\citep{Tapia:2013kf,Lopez:2014en}, in part simply because there are more physical virions per infectious unit (\FIG{viruspopulations}A,B).
We confirmed that our viral stocks induced less interferon per infectious unit than a stock propagated at higher MOI (\FIG{viruspopulations}C).

\begin{figure}
\centerline{\includegraphics[width=0.7\linewidth]{figures/Validating_barcode_virus/validating_populations_D02.pdf}}
\caption{\label{fig:viruspopulations} The viral stocks in our experiments are relatively pure of defective particles. 
{\bf (A)}
Our viral stocks have a higher ratio of infectious particles to HA virion RNA compared to a high-defective stock propagated at high MOI.
HA viral RNA was quantified by qPCR on virions. 
Error bars $\pm$ S.D., n=6 (qPCR replicates). 
{\bf (B)} 
Our viral stocks have a higher ratio of infectious particles to particles capable of expressing HA protein.
A549 cells were infected at an MOI of 0.1, and the percentage of cells expressing HA protein at 9 hours post-infection was quantified by antibody staining and flow cytometry.
{\bf (C)} 
Our viral stocks are less immunostimulatory than virus propagated at high MOI when used at the same number of infectious units as calculated by TCID50.
Note that this fact does not necessarily imply that they are more immunostimulatory per virion, as the high-MOI stocks also have more virions per infectious unit as shown in the first two panels. 
Measurements of \textit{IFNB1} transcript by qPCR normalized to the housekeeping gene L32 in A549 cells at 10 hours post infection at an MOI of 0.5.
Error bars $\pm$ S.D., n=3.
Note that MOIs were calculated by TCID50 on MDCK-SIAT1 cells, whereas the experiments in this figure involved infection of A549 cells.
}
\figsupp[Full flow cytometry data for panel B . \label{figsupp:HAstain}]{
Full flow cytometry data for  \FIG{viruspopulations}B. 
A549 cells were infected at an MOI of 0.1 as calculated by TCID50 on MDCK-SIAT1 cells. 
{\bf (A)} Uninfected gating control.
{\bf (B)} Cells infected with the wild-type virus stock used in our experiments.
{\bf (C)} Cells infected with synonymously barcoded virus stock used in our experiments.
{\bf (D)} Cells infected with a stock of wild-type virus propagated at a high MOI, and therefore enriched in defective particles.
}{\includegraphics[width=0.7\linewidth]{figures/Validating_barcode_virus/HA_stain_supplement.png}}
\end{figure}

@@

We recovered between 3,000 and 4,000 cells for each sample (\FIG{cells}A). 
As expected for a low-MOI infection, most cells expressed little or no viral mRNA (\FIG{cells}B, \FIGSUPP[cells]{viral_mRNA_cumulfrac}).
Also as expected, the amount of viral mRNA per cell among infected cells increased over time (\FIG{cells}B, \FIGSUPP[cells]{viral_mRNA_cumulfrac}).
But what was most notable was how widely the number of viral mRNA molecules varied among infected cells.
While the fraction of mRNA derived from virus was $<$0.1\% for most cells, viral mRNA constituted half the transcriptome in a few cells at 8 and 10 hours (\FIG{cells}B, \FIGSUPP[cells]{viral_mRNA_cumulfrac}).

\begin{figure}
\centerline{\includegraphics[width=\linewidth]{figures/p_cell_mRNA_summary.pdf}}
\caption{\label{fig:cells}
There is a very wide distribution in the amount of viral mRNA per cell.
{\bf (A)} 
Number of cells sequenced for each sample.
{\bf (B)} 
The number of cellular and viral mRNAs detected for each cell is plotted as a point.
The blue lines show the overall distribution of the number of cellular mRNAs per cell.
The orange rug plot at the left of each panel shows the distribution of the number of viral mRNAs per cell.
Cells outside the dotted green lines were considered outliers with suspiciously low or high amounts of cellular mRNA (possibly derived from two cells per droplet), and were excluded from all subsequent analyses.
\FIGSUPP[cells]{viral_mRNA_cumulfrac} shows the exact distributions of the fraction of viral mRNA per cell.
}
\figsupp[Cumulative fraction plot of proportion of total mRNA from virus. \label{figsupp:viral_mRNA_cumulfrac}]{
For each sample, this plot shows the fraction of all cells that derive at least the indication fraction of their mRNA from influenza virus.
}{\includegraphics[width=0.5\linewidth]{figures/p_mRNA_per_cell_cumul.pdf}}
\end{figure}

A complicating factor is that uninfected cells could have small amounts of viral mRNA due to leakage of transcripts from lysed cells.
It is therefore important to establish a threshold for identifying truly infected cells.
We can do this by taking advantage of the fact that roughly half the infecting virions bear synonymous barcodes.
Reads derived from lysed cells will be drawn from both wild-type and synonymously barcoded viral transcripts.
However, most cells are infected by at most one virion, and so the reads from truly infected cells will usually derive almost entirely from one of the two viral variants. 
\FIG{viralbarcodes}A shows the fraction of viral reads in individual cells from each viral variant, and \FIG{viralbarcodes}B indicates the fraction of viral reads from the most abundant variant in that cell.
Most cells with large amounts of viral mRNA have viral transcripts exclusively derived from one viral variant -- indicating non-random partitioning as expected from viral infection.
However, cells with a small amount of viral mRNA often have viral transcripts from both variants, as expected from the random partitioning associated with simple mRNA leakage.
Finally, a few cells with large amounts of viral mRNA have viral transcripts from both variants, likely reflecting co-infection.

\begin{figure}[t!]
\centerline{\includegraphics[width=0.82\linewidth]{figures/p_frac_flu_summary.pdf}}
\caption{\label{fig:viralbarcodes}
Synonymous barcodes on the viral mRNAs distinguish true infections from cells that contain viral mRNAs derived from leakage of lysed cells.
{\bf (A)}
Cells with at least two viral mRNAs for which the barcode could be called, arranged in order of increasing influenza transcript counts.
Bar heights denote the number viral mRNAs on a log\textsubscript{10} scale, bar coloring is linearly proportional to the fractions of viral mRNAs derived from wild-type and synonymously barcoded virus.
{\bf (B)}
Same as (A), but each bar is colored according to the relative fraction of the more common (major) and less common (minor) virus variant.
At low levels of viral mRNA there is often a roughly equal mix, suggesting contamination with viral mRNAs leaked from lysed cells.
At higher levels of viral mRNA, cells generally have only one viral variant, suggesting infection initiated by a single virion.
A few cells are also obviously co-infected with both viral variants.
{\bf (C)}
We determined a threshold for calling ``true'' infections by finding the amount of viral mRNA per cell at which the viral barcode purity no longer increases with more viral mRNA.
The purity is the fraction of all viral mRNA in a cell derived from the most abundant viral barcode in that cell.
We fit a curve (orange line) to the mean purity of all cells with more than the indicated amount of viral mRNA, and drew the cutoff (dotted green line) at the point where this curve stopped increasing with the fraction of total mRNA derived from virus.
This plot illustrates the process for the 10-hour sample, see \FIGSUPP[viralbarcodes]{thresholds} for similar plots for other samples.
See the Methods for details.
{\bf (D)}
The number of cells identified as infected and co-infected for each sample, as well as the number of cells with any viral read.
For all subsequent analyses, we subsampled the number of uninfected cells per sample to the greater of 50 or the number of infected cells.
{\bf (E)} 
Distribution of the fraction of mRNA per cell derived from virus for both infected and co-infected cells.
\FIGSUPP[viralbarcodes]{cumulfracflu} shows these same data in a cumulative fraction plots and calculates Gini coefficients to quantify the heterogeneity in viral mRNA load.
}

@@

\figsupp[Thresholds for calling infected cells. \label{figsupp:thresholds}]{
Cell lysis can lead cells to the spurious association of small amounts of extraneous mRNA with individual cells.
We wanted to avoid classifying as infected cells that had simply acquired such lysis-derived viral mRNA.
The amount of lysis-derived viral mRNA will vary among samples as a function of both the lysis rate during the cell preparation (which always varies slightly from sample to sample in the 10X procedure) and with the amount of total viral mRNA for that sample (the more viral mRNA, the more there is to be acquired from lysed cells).
As is shown in \FIG{viralbarcodes}B, the 8hr-2 and 10hr sample clearly have an enrichment of mixed barcodes in cells with small numbers of viral mRNA.
For each sample, we calculated the mean purity of all cells with at least the indicated amount of viral mRNA, and determined the threshold amount of viral mRNA where purity no longer increased by finding the first maxima in a loess curve fit (orange line).
We called the threshold at this point of maximum purity (dotted green line).
For the 6hr and 8hr samples there is no indication of contamination from lysis-derived reads, as \FIG{viralbarcodes}B shows no increase in mixed barcodes in low viral mRNA cells.
Therefore, for these samples we simply set a threshold of requiring at least 2$\times$10$^{-4}$ of the total mRNA to come from virus, which corresponds to $\sim$2 viral mRNAs for the typical cell with $10^4$ total reads (\FIG{cells}B).
}{\includegraphics[width=\linewidth]{figures/p_purity_threshold.pdf}}

@@

We determined the threshold amount of viral mRNA per cell for each sample at which the barcode partitioning clearly resulted from infection rather than leakage (\FIG{viralbarcodes}C, \FIGSUPP[viralbarcodes]{thresholds}), and used these thresholds to annotate cells that we were confident were truly infected.
We also annotated as co-infected cells above this threshold that had mRNA from both viral variants.
\FIG{viralbarcodes}D shows the number of cells annotated as infected and co-infected for each sample -- these cells are just a small fraction of the number of cells with any viral read.
These annotation thresholds are conservative, and may miss some true low-level infections.
However, it is important that the analyses below are restricted to cells that are truly infected with virus, so we accepted the possible loss of some low-level infections in order to avoid false positives.
In addition, the synonymous viral barcodes only identify co-infections by viruses with different barcodes -- since the barcodes are at roughly equal proportion, we expect to miss about half of the co-infections.
Since we annotate about $\sim$10\% of the infected cells as co-infected by viruses with different barcodes (\FIG{viralbarcodes}D), we expect another $\sim$10\% of the infected cells to also be co-infected but not annotated as so by our approach.
Because most cells are not infected, we subsampled the uninfected cells to the numbers shown in \FIG{viralbarcodes}D to balance the proportions of infected and uninfected cells for all subsequent analyses.

Strikingly, the extreme variation in the number of viral transcripts per cell remains even after we apply these rigorous criteria for annotating infected cells (\FIG{viralbarcodes}E). 
The fraction of viral mRNA per infected cell follows a roughly exponential distribution, with many cells having few viral transcripts and a few cells having many.
At 6 and 8 hours $<$10\% of infected cells are responsible for over half the viral transcripts, while at 10 hours $<$15\% of infected cells produce over half the viral transcripts (\FIGSUPP[viralbarcodes]{cumulfracflu}).
One way to quantify the heterogeneity of a distribution is to calculate the Gini coefficient~\citep{gini1921measurement}, which ranges from 0 for a completely uniform distribution, to 1 for a maximally skewed distribution.
\FIGSUPP[viralbarcodes]{cumulfracflu} shows the Gini coefficients for the distribution of viral mRNA across infected cells for each sample.
The Gini coefficients are $\ge$0.64 for all samples.
As a fun point of comparison, these Gini coefficients indicate that the distribution of viral mRNA across infected cells is more uneven than the distribution of income in the United States~\citep{alvaredo2011note}.

One possible source of heterogeneity in the amount of viral mRNA per cell is variability in the timing of infection.
If some cells are infected earlier in the experiment than others, then they might have substantially more viral mRNA.
However, several lines of evidence indicate that this is not the major cause of heterogeneity across cells.
First, the sample for which the infection inoculum was never removed (8hr-2) only shows slightly more heterogeneity than samples for which the inoculum was washed away after one hour (\FIG{viralbarcodes}E, \FIGSUPP[viralbarcodes]{cumulfracflu}), despite the fact that the potential time window for infection is much longer in the former sample.
Second, in an independent experiment, we performed completely synchronized infections by pre-binding virus to cells on ice and then washing away unbound virus before bringing the cells to 37$^{\circ}$C~\jdbcomment{Alistair, can you add relevant citation for this method?}.
As shown in \FIGSUPP[viralbarcodes]{synchronize}, flow cytometry staining found that the heterogeneity in the levels of individual viral proteins was not markedly different for these synchronized infections than in the absence of pre-binding and washing.
Finally, viral mRNA expression from the secondary spread of virus from infected cells does not appreciably occur during the timeframes of our experiments, since \FIG{viralbarcodes}B does not show the pervasive presence of mixed barcodes that would occur in this case.
Therefore, variability in the timing of infection is not the dominant cause of the cell-to-cell heterogeneity in our experiments.

Notably, \FIG{viralbarcodes}E shows that there are co-infected cells with both low and high amounts of viral mRNA, suggesting that the initial infectious dose does not drive a simple continuous increase in viral transcript production.
In support of this view, we used flow cytometry to quantify the levels of individual viral proteins in cells infected at various MOIs or for which we could delineate co-infection status (\FIGSUPP[viralbarcodes]{flowdose}).
This analysis shows that sub-populations of cells that express similarly low and high levels of viral proteins persist across a wide range of infectious doses, although co-infection can influence the relative proportion of infected cells that fall into these sub-populations (\FIGSUPP[viralbarcodes]{flowdose}).
